---
title: "Adaptation to changing environments across real geographies"
author: "Peter Ralph"
date: "January 13, 2016"
---

```{r setup, include=FALSE}
knitr::opts_knit$set(self.contained=TRUE)
fig.dim <- 8
knitr::opts_chunk$set(
           fig.width=1.4*fig.dim,
           fig.height=fig.dim,
           fig.align='center',
           dev.args=list(pointsize=24),
           error=FALSE )
```



Climate change
--------------

*Rapid* climate change:

- occurring over tens of generations
- different environment $\Rightarrow$ strong selective pressure

Ecological dynamics important,
but today, just thinking about the population genetics.


Questions
---------

1.  Can species adapt? (maintaining something like their present range)

2.  If so, how will genetic diversity be affected?

3.  How will independently arisen adaptations interact?


--------------

1.  Is there existing variation for the selected trait(s)?

    - genetic load?
    - due to local adaptation?
    - temporal fluctuations?
    - balancing selection?

2.  What is the genetic architecture of the trait?

    - mutational target size?
    - interactions?
    - dominance?
    - pleiotropy?

3.  Can adaptations spread?

    - e.g., from south to north
    - or from locations of existing variants


Population genetics parameters
------------------------------

1.  population size

    - availability of standing deleterious variation
    - influx of new mutations

2.  strength of selection

    - probability of establishment of rare variants
    - rate of spread of adaptive variants

3.  rate of genetic drift

    - rate of local loss of variation
    - probability of establishment of rare variants


Geography affects
-----------------

(Really, geography ${}+{}$ dispersal ${}+{}$ habitat dynamics.)

- Rate of drift (inbreeding)

- Speed of spread

- Pre-existing adaptations



Probability of establishment
----------------------------

The *probability of establishment* 
in a large population is roughly
$$ p_\text{estab} \approx 2 s / \nu . $$

If $s < 1/N_\text{loc}$, *local inbreeding* 
may reduce $p_\text{estab}$.


Rate of drift
-------------

$$ N_\text{loc} \propto \frac{ \#\{ \text{ individuals* within 1$\sigma$ } \} }{ \nu }, $$

where 

-  $\sigma$ is the **dispersal distance**
-  *individuals* are haploid,
-  $\nu = N/N_e$, the strength of drift.


-----------------

```{r setup_habitat_one_sigma, echo=FALSE, cache=2}
library(raster); library(rgeos); library(landsim)
if (!file.exists("desert_veg_central_small.RData")) {
    layer.dir <- "/home/peter/projects/patchy-landscapes/layers"
    vegmap <- readOGR(file.path(layer.dir,"desert_veg/central","ds166"))
    this.ex <- c(354660.90906967, 381245.239994035, -273140.670781638, -251867.619631655)
    smallveg <- crop(vegmap,this.ex)
    source(file.path(layer.dir,"layer_utils.R",chdir=TRUE))
    counties <- get_counties(smallveg)
    elev <- get_dem(smallveg)
    shade <- get_elev(smallveg)
    smallveg$LABEL_1 <- droplevels(smallveg$LABEL_1)
    save(smallveg,counties,elev,shade,file="desert_veg_central_small.RData")
} else {
    load("desert_veg_central_small.RData")
}

hcols <- c( Blackbrush="#FF9400FF", Creosote="#FFC500FF", 'Creosote-Brittlebush'="#FFF700FF", 'Desert Holly'="#D6FF00FF", 'Joshua Tree'="#00FF52FF", 'Low Elevation Wash System'="#00E6FFFF", 'Mid Elevation Wash System'="#0052FFFF", 'Mojave Yucca'="#1000FFFF" )
hcols <- hcols[match(levels(smallveg$LABEL_1),names(hcols))]
```
```{r habitat_one_sigma, echo=FALSE, cache=2, depends="setup_habitat_one_sigma"}

refpoint.list <- list( 
       "Creosote" = cbind(x=369276.5, y=-260411.1),
       "Low Elevation Wash System" = cbind(x=365959.7, y=-260357.6)
   )
refcol <- c( Creosote="red", "Low Elevation Wash System"="blue" )
radii <- c(1000,5000)

layout(t(1:2))
par(mar=c(0,0,0,0)+.5)
for (vegtype in names(refpoint.list)) {
    thisveg <- subset( smallveg, LABEL_1==vegtype )
    # refpoints <- spsample( thisveg, n=1, type='random' )
    refpoints <- SpatialPoints( refpoint.list[vegtype], proj4string=CRS(proj4string(smallveg)) )
    refcircles <- make_circles( refpoints[rep(1,length(radii))], radii=radii )
    habcircles <- gIntersection( refcircles, crop(thisveg,3*extent(refcircles)), byid=c(TRUE,FALSE) )

    plot( smallveg, col=adjustcolor(hcols[smallveg$LABEL_1],0.2) ) #, add=TRUE )
    plot( habcircles, col=adjustcolor(refcol[vegtype],0.5), add=TRUE, border=NA )
    lines(refcircles, lty=2 )
    points(refpoints,pch="*", col='red')
}

```

Speed of spread
---------------

Alleles that are advantageous when rare
spread, like *wildfire*. 
(a "pushed" wave)

```{r setup_wildfire_spread, echo=FALSE, cache=2}
if (!file.exists("desert_veg_central_small_raster.RData")) {
    z <- lapply( levels(smallveg$LABEL_1), function (vt) { gUnaryUnion( subset(smallveg,LABEL_1==vt) ) } )
    rz <- lapply(z, function (x) { rasterize(x, raster(extent(smallveg),res=100), getCover=TRUE ) } )
    rasterveg <- do.call( stack, rz )
    # rasterveg <- rasterize(smallveg,raster(extent(smallveg),res=50),"LABEL_1",getCover=TRUE)
    values(rasterveg) <- values(rasterveg)/100
    names(rasterveg) <- levels(smallveg$LABEL_1)
    save(rasterveg, file="desert_veg_central_small_raster.RData")
} else {
    load("desert_veg_central_small_raster.RData")
}
```
```{r setup_demog, echo=FALSE, cache=2}
demog <- demography(
        prob.seed = 0.2, fecundity = 100,
        prob.germination = vital( 
                 function (N, ...) {
                     out <- r0 / ( 1 + rowSums(N)/carrying.capacity )
                     return( cbind( aa=out, aA=s*out, AA=s^2*out ) )
                 },
                 r0 = 0.01,  s = 1.5 ),
        prob.survival = 0.9,
        pollen.migration = migration(
                 kern = function (x) { exp(-sqrt(x)) },
                 sigma = 300, radius = 1200, normalize = NULL ),
        seed.migration = migration(
                 kern = "gaussian", sigma = 100,
                 radius = 1200, normalize = 1 ),
        genotypes = c("aa","aA","AA")
    )
```
```{r run_sim, echo=FALSE, cache=2, depends=c("setup_wildfire_spread","setup_habitat_one_sigma","setup_demog")}
if (!file.exists("wildfire-sims.RData")) {
    wildfire.vegtypes <- c("Low Elevation Wash System","Creosote")
    mutloc <- list( center=SpatialPoints( cbind(x=358709.2, y=-258082.9), proj4string=CRS(proj4string(smallveg)) ) )
    wildfire.sims <- lapply( wildfire.vegtypes,  function (vegtype) {
            habitat <- rasterveg[[make.names(vegtype)]]
            carrying.capacity <- 120 * prod(res(habitat))/270^2  # estimated from google maps near desert queen mine
            habitable <- (!is.na(values(habitat)) & values(habitat)>0)
            pop <- population( 
                              habitat = habitat,
                              # accessible = !is.na(values(habitat)),
                              accessible = habitable,
                              habitable = habitable,
                              genotypes = c("aa","aA","AA"),
                              carrying.capacity = carrying.capacity,
                              N = cbind( aa=rpois(sum(habitable),carrying.capacity),
                                         aA=0, AA=0 )
                         )
            demog <- setup_demography( demog, pop )
            # summary functions
            # distance
            dists <- pointDistance( mutloc$center, xyFromCell(pop$habitat,which(pop$habitable)), lonlat=FALSE )
            # distance to the mutation center for each habitable cell
            volume_fun <- function (N) { colSums(N>0) }
            radius_fun <- function (N) { apply( N, 2, function (x) { suppressWarnings( max( dists[ x>0 ] ) ) } ) }
            summary.funs <- list( totals=function(N){colSums(N)}, volume=volume_fun, radius=radius_fun)
            # burn-in
            sim <- simulate_pop( pop, demog, times=c(0,100) )
            pop$N[] <- sim$N[,,2]
            mutloc$cell.number <- match( cellFromXY( habitat, mutloc$center ), which(habitable) )
            for (ntries in 1:100) {
                pop <- set_N( pop, i=which(pop$habitable), j="aA", 
                         value=ifelse(1:nrow(pop$N)==mutloc$cell.number,3,0) )
                sim <- simulate_pop( pop, demog, times=seq(0,100,length.out=11),
                                 summaries= summary.funs )
                total.aA <- sim$summaries[[1]][nrow(sim$summaries[[1]]),"aA"]
                if ( total.aA>0 ) { break }
            }
            sim <- extend_simulation( sim, pop, demog, times=seq(sim$t,1000,length.out=101), summaries=summary.funs )
            return(list(pop=pop,sim=sim))
        } )
    names(wildfire.sims) <- wildfire.vegtypes
    save(mutloc,wildfire.sims,wildfire.vegtypes, file="wildfire-sims.RData")
} else {
    load("wildfire-sims.RData")
}
```
```{r animate_wildfire, echo=FALSE, cache=2, depends="run_sim"}
for (vegtype in names(wildfire.sims)) {
    outfile <- paste("spatial-sweep-",gsub(" ","_",vegtype),".mp4",sep='')
    if (!file.exists(outfile)) {
        plot.simulation( wildfire.sims[[vegtype]]$sim, wildfire.sims[[vegtype]]$pop,
                    animate=outfile, duration=20, width=12, height=4, pointsize=18 )
    }
}
```

------------------

![Wash Systems in Central Mojave](spatial-sweep-Low_Elevation_Wash_System.mp4){ width=10in height=3.33333in }

------------------

![Creosote Scrub in Central Mojave](spatial-sweep-Creosote.mp4){ width=10in height=3.33333in }

------------------

```{r show_speeds_wash, cache=2, depends="run_sim", echo=FALSE}
vegtype <- "Low Elevation Wash System"
speeds <- list()
speeds[[vegtype]] <- with( list2env(wildfire.sims[[vegtype]]), {
        matplot( sim$summary.times, sim$summaries[["radius"]][,c("aA","AA")], type='l', lty=1:2, col=hcols[vegtype],
                xlab="generation", main='radius', ylab='distance (m)', xlim=c(0,500) )
        legend( "bottomright", legend=c(pop$genotypes[-1],names(wildfire.sims)), lty=c(1:2,1,1), 
               col=c("black","black",hcols[names(wildfire.sims)]), cex=0.75 )
        AA.lims <- c(1.2,0.8)*range(sim$summaries[["radius"]][,"AA"],finite=TRUE)
        AA.usethese <- which( ( sim$summaries[["radius"]][,"AA"] > AA.lims[1] ) 
                              & ( sim$summaries[["radius"]][,"AA"] < AA.lims[2] ) )
        AA.speed <- diff( sim$summaries[["radius"]][range(AA.usethese),"AA"] ) / diff( sim$summary.times[range(AA.usethese)] )
        if (sum(!is.na(AA.usethese))>0) {
            points(  sim$summary.times[range(AA.usethese)], sim$summaries[["radius"]][range(AA.usethese),"AA"], pch=20, col=hcols[vegtype] )
            abline( sim$summaries[["radius"]][min(AA.usethese),"AA"]-AA.speed*sim$summary.times[min(AA.usethese)], AA.speed, col=hcols[vegtype], lty=3 )
        }
        AA.speed
    } )
vegtype <- "Creosote"
speeds[[vegtype]] <- with( list2env(wildfire.sims[[vegtype]]), {
        matlines( sim$summary.times, sim$summaries[["radius"]][,c("aA","AA")], type='l', lty=1:2, col=hcols[vegtype] )
        AA.lims <- c(1.2,0.8)*range(sim$summaries[["radius"]][,"AA"],finite=TRUE)
        AA.usethese <- which( ( sim$summaries[["radius"]][,"AA"] > AA.lims[1] ) 
                              & ( sim$summaries[["radius"]][,"AA"] < AA.lims[2] ) )
        AA.speed <- diff( sim$summaries[["radius"]][range(AA.usethese),"AA"] ) / diff( sim$summary.times[range(AA.usethese)] )
        if (sum(!is.na(AA.usethese))>0) {
            points(  sim$summary.times[range(AA.usethese)], sim$summaries[["radius"]][range(AA.usethese),"AA"], pch=20, col=hcols[vegtype] )
            abline( sim$summaries[["radius"]][min(AA.usethese),"AA"]-AA.speed*sim$summary.times[min(AA.usethese)], AA.speed, col=hcols[vegtype], lty=3 )
        }
        AA.speed
    } )
# speeds
```


------------------

**Continuous habitats**

- speed: $\sigma \sqrt{2s}$ (Fisher/KPP)
    $s$ is *growth rate when rare*

- leptokurtic dispersal: patchy growth, accelerates

    **Oskar**

- slowed by drift and 
    - *tortuosity*: ratio of length of curve to distance between its ends

    **Nelson**

-------------------

**Patchy habitats**

- spread governed by rate of migrants (sparks)
    ${} \times p_\text{estab}$

- if growth rate in gaps is $1-m$,
    rate of migrants $\propto e^{-x m/\sigma}$

    **pgen**

- if gaps are totally uninhabitable, depends on *long distance dispersal*


Both depend critically on *dispersal* mechanism.



Mojave habitats: Vegetation classification
---------------

**image**


Mojave habitats: Environmental niche models
---------------

- Joshua tree
- others


-----------------

**Proportion of neighborhood occupied:**

```{r vol_versus_radius, echo=FALSE, cache=2, fig.width=1.8*fig.dim}
source("habitat-shapes.R")
mean.areas <-  sapply( central.mojave.area.results, colMeans )
area.units <- 1e4 # ha
layout(t(1:2),width=c(3,1))
par(mar=c(4,4,1,1)+.1)
matplot(central.mojave.radii, area.units*mean.areas/(pi*central.mojave.radii^2), type='l', log='x', ylab="proportion of circle covered", xlab="radius (m)")
opar <- par(mar=c(par("mar"),0)[c(1,5,3,5)])
plot(0,xaxt='n',yaxt='n',xlab='',ylab='',type='n',bty='n')
legend("topleft",legend=colnames(mean.areas),lty=1:5,col=1:6, cex=0.6, bty='n')
```

---------------

**Speed of spread:**

```{r habitat_speeds, echo=FALSE, cache=2}
# provides speed.results:  a list, each component has
#     svals : selection coefficients
#     sumtabs : list of the same length as svals giving the results of a simulation with the corresponding selection coefficient
source("habitat-speeds.R")
speedfun <- function (sumtab) {
    # returns (x0,x1,y0,y1,(y1-y0)/(x1-x0))
    AA.lims <- c(1.2,0.8)*range(sumtab[,"radius_AA"],finite=TRUE)
    AA.usethese <- which( ( sumtab[,"radius_AA"] > AA.lims[1] ) 
                          & ( sumtab[,"radius_AA"] < AA.lims[2] ) )
    out <- if (length(AA.usethese)>1) {
        AA.speed <- diff( sumtab[range(AA.usethese),"radius_AA"] ) / diff( sumtab[range(AA.usethese),"t"] )
        c(  sumtab[range(AA.usethese),"t"], sumtab[range(AA.usethese),"radius_AA"], AA.speed )
    } else { rep(NA,5) }
    names(out) <- c("x0","x1","y0","y1","speed")
    out
}
speeds <- lapply( speed.results, function (ss) { cbind( s=ss$svals, t(sapply( ss$sumtabs, speedfun )) ) } )
speeds <- data.frame( habitat=rep(names(speeds),sapply(speeds,nrow)), do.call( rbind, speeds ) )
```

<!-- Naive prediction has $\sigma=250$. -->
```{r speed.summary, echo=FALSE}
par(mar=c(4,4,1,1)+.1)
with( speeds, plot( jitter(s), speed,col=habitat, xlab="s", ylab="speed", pch=20 ) )
with( speeds, matlines( sort(unique(speeds$s)), with( speeds, tapply( speed, list(s,habitat), mean, na.rm=TRUE ) ), lty=1 ) )
legend("topleft",pch=20,col=seq_along(levels(speeds$habitat)), legend=levels(speeds$habitat), cex=0.75 )
```

----------------------------

**Probability of establishment:**
```{r plot_pestab, echo=FALSE}
source("establishment-probs.R") # provides establishment.probs
pestabs <- data.frame( habitat=rep(names(establishment.probs),sapply(establishment.probs,nrow)),
        do.call( rbind, establishment.probs ) )
plot(pestabs$s, pestabs$p_estab, col=pestabs$habitat, xlab="s", ylab="prob of establishment", pch=20)
legend("topleft",pch=20,legend=levels(pestabs$habitat),cex=0.75,col=1:nlevels(pestabs$habitat))
for (k in seq_along(establishment.probs)) {
    abline(0,coef(lm(p_estab ~ s + 0, data=data.frame(establishment.probs[[k]]))), col=k)
}
```



Conclusions
-----------

Wildly varying predictions, depending on:

- genetic architecture
- population density
- what is "habitat"
- long-distance dispersal

What's clear:

- fast adaptation can't be from single, new mutations
- truly isolated pockets are on their own (and have been already)


Thanks!
-------

- Erik Lundgren (USC)
- Julie Evens (CNPS)
- Greg Suba (CNPS)
- R's GIS tools: `sp`, `raster`, `rgeos`, ...

An R package: github.com/petrelharp/landsim

